Community Detection Methods for the Generalized Random Dot Product Graph Model

### 2021-05-24 proposal/slides.Rmd
updated slides

Definition based on Noroozi, Rimal, and Pensky (2020).

$A \sim PABM(\{\lambda^{(kl)}\}_K)$ iff

1. w.l.o.g., organize $P$ such that each block 
$P^{(kl)} \in [0, 1]^{n_k \times n_l}$ contains edge 
probabilities between communities $k$ and $l$.
2. Organize parameters as vectors such that 
$\lambda^{(kl)} \in \mathbb{R}^{n_k}$ 
are the popularity parameters of members of community $k$ to community $l$.  
$\{\lambda^{(kl)}\}_K$ is the set of $K^2$ popularity vectors.
3. Then we can write each block of $P$ as 
$P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$.
4. Sample $A \sim BernoulliGraph(P)$.

@@

**Theorem** (KTT): $A \sim PABM(\{\lambda^{(kl)}\}_2)$ is equivalent to 
$A \sim GRDPG_{3, 1}(X U)$ for block diagonal $X$ 
constructed from $\{\lambda^{(kl)}\}_2$ and $U \in \mathbb{O}(4)$.

@@

$$Y = \begin{bmatrix}
\lambda^{(11)} & 0 & \lambda^{(12)} & 0 \\
0 & \lambda^{(21)} & 0 & \lambda^{(22)}
\end{bmatrix}$$

$$P = X Y^\top = \begin{bmatrix}
\lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
\lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)} (\lambda^{(22)})^\top
\end{bmatrix}$$

Proof (cont'd):

$$Y = X \Pi$$

$$\Pi = \begin{bmatrix} 
1 & 0 & 0 & 0 \\ 
0 & 0 & 1 & 0 \\
0 & 1 & 0 & 0 \\
0 & 0 & 0 & 1
\end{bmatrix} = 
U I_{3, 1} U^\top$$

@@

$$P = (X U) I_{3, 1} (X U)^\top$$

@@

* $p = K (K + 1) / 2$
* $q = K (K - 1) / 2$
* $U \in \mathbb{O}(K^2)$
* $X$ is block diagonal and composed of $\{\lambda^{(kl)}\}_K$ 
with each block corresponding to a community  
$\implies$ if $x_i^\top$ and $x_j^\top$ are two rows of $X U$ corresponding 
to different communities, then $x_i^\top x_j = 0$.

>>>>

**Def** Popularity Adjusted Block Model (Sengupta and Chen, 2017):

Let each vertex $i \in [n]$ have $K$ popularity parameters
$\lambda_{i1}, ..., \lambda_{iK} \in [0, 1]$. 
Then $A \sim PABM(P)$ if each $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$.

**Lemma** (Noroozi, Rimal, and Pensky, 2020):

$A$ is sampled from a PABM if 
$A \sim BernoulliGraph(P)$ such that:

1. Let $P^{(kl)}$ denote the $n_k \times n_l$ matrix of edge probabilities 
between communities $k$ and $l$. 
2. Organize popularity parameters as vectors 
$\lambda^{(kl)} \in \mathbb{R}^{n_k}$ 
such that $\lambda^{(kl)}_i$ is the popularity parameter 
of the $i$^th^ vertex of community $k$ towards community $l$. 
3. Each block can be decomposed as 
$P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$.

**Notation**: $A \sim PABM(\{\lambda^{(kl)}\}_K)$.

@@

**Theorem** (KTT): 
Given popularity vectors $\{\lambda^{(kl)}\}_2$, we can define block diagonal $X \in \mathbb{R}^{n \times 4}$ and $U \in \mathbb{O}(4)$ such that $A \sim PABM(\lambda^{(kl)}_2)$ is equivalent to $A \sim GRDPG(X U)$.

@@

@@

$$P = (X U) I_{3, 1} (X U)^\top = 
\begin{bmatrix}
\lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
\lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)} (\lambda^{(22)})^\top
\end{bmatrix}$$

@@

* $p = K (K + 1) / 2$, $q = K (K - 1) / 2$
* $U \in \mathbb{O}(K^2)$
* $X \in \mathbb{R}^{n \times K^2}$ is block diagonal and 
composed of $\{\lambda^{(kl)}\}_K$ 
with each block corresponding to a community  
$\implies$ if $x_i^\top$ and $x_j^\top$ are two rows of $X U$ corresponding 
to different communities, then $x_i^\top x_j = 0$.
